Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands

### 2024-12-07 index.qmd
k

This study develops a technique for estimating BFI in ungauged basins in Arizona and determines the state’s long-term mean BFI from 1991 to 2020. Due to interannual variability in BFI, a 30-year period of record is used to evaluate the characteristic base-flow dynamics of Arizona. Using a machine learning model trained with the hydrogeologic characteristics of surface water basins, we first determine annual BFIs for each year in the 1991-2020 study period for ungauged river reaches and physiographic regions of Arizona. Long-term mean BFI data for the study period are then estimated. Regional patterns and trends in base flow of instrumented sites are subsequently investigated. These results provide novel analyses of low flow processes in Arizona...######

@@

The state of Arizona, located in the southwest United States, covers a total area of 295,253 km$^2$. Arizona is divided into two primary physiographic provinces: the Colorado Plateau in the northeast, and the Basin-and-Range region in the northwest and south. The Central Highlands is a transition zone consisting of scattered basins separated by the mountainous foothills of the Mogollon Rim. The Colorado Plateau is dominated by high-elevation desert with an average elevation of 1,936 masl (6,352 ft). Average temperatures range from -6$^\circ$C (20$^\circ$F) to 26$^\circ$C (80$^\circ$F) and it averages 580 mm (23 in) of precipitation. The Basin-and-Range region has a semi-arid to arid climate with an average elevation of 490 masl (1600 ft). The average precipitation is 200 mm (8 in) annually with average temperatures ranging from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F).

@@

Arizona's hydrology varies seasonally and spatially between its physiographic regions. In the summer, localized and intense convective precipitation events are driven by the North American Monsoon, while in the winter, orographic precipitation comes from Pacific frontal systems [@eastoe2019mtnblock]. While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002]. As such, 94% of streams in Arizona are ephemeral or intermittent [@epa_AZephemeral]. Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack. While winter precipitation provides only 30% of annual averages, it provides the bulk of water for natural reservoirs [@sheppard2002].

@@

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. @odonnell2016 found that 8-10 years of calibration data were necessary to account for climate variability in paired watershed studies in the region. Following this criteria, a minimum of 10 years of record was chosen. Any years missing more than 30 days of streamflow were removed from the possible record period. This analysis focuses on natural, streamflow-influencing recharge, excluding streamgages affected by regulation or diversions. These gages were identified through annual reports on water data published by the USGS [@usgs2010]. Finally, streamgages on the Colorado River were excluded from the dataset on the basis that they represent managed flows shared by Colorado River Compact states. Following these parameters, 205 USGS streamgages with acceptable periods of record were used in this study (@fig-study-area).

Watersheds across the United States are delineated by the USGS using a hydrologically-defined network. This system delineates the country using hierarchical hydrologic unit codes (HUCs), where each subsequent basin includes the digits of the enclosing basin. Here, 8-digit HUCs (HUC 8s) are used to divide Arizona into 84 sub-basins that are fully or partially in the state (@fig-study-area). These HUC 8 sub-basins are analogous to medium-sized river basins and are defined by surface water characteristics. \######

The annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]). The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM]**.** Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance. Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].

@@

A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects. Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins. By calculating these percentages, we aimed to get a more comprehensive understanding of landscape composition across space. Land cover from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology/karst from USGS were all similarly averaged across the basins. Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.

@@

Base flow was calculated using a single-parameter, recursive digital filter technique from [@nathan1990]. This base-flow separation technique is based on a recursive digital filter used in signal analysis that separates high-frequency signals (quickflow) from low-frequency signals (base flow) [@lyne1979]. @eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study. This technique has been used in [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018]. It takes the form of

@@

Statisical analyses were performed on mean annual BFI values of instrumented streamgages. We examined the presence of temporal trends in mean annual BFI using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric]. The Mann-Kendall test was used to determine trends in BFI and trend significance of gauged reaches.

These trends were produced using a Mann-Kendall test for monotonic trends. This test expects that the data is non-parametric and that the observations are not autocorrelated over time. Observations of BFI at 4 streamgages show significant autocorrelation, though only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a significant trend in BFI. Due to this autocorrelation, the streamgage was removed from the trend analysis as the autocorrelation may inflate the variance of the Mann-Kendall statistic leading to an over/underestimate of the trend (CITATION).

>>>>

This study develops a technique for estimating BFI in ungauged basins in Arizona and determines the state’s long-term mean BFI. Using a machine learning model trained with the hydrogeologic characteristics of surface water basins, we estimate mean BFI in ungauged basins over a period of record from 1991-2020. Regional patterns and trends in base flow and BFI of instrumented sites are subsequently investigated. These results provide novel analyses of low flow processes in Arizona, furthering our understanding of climate controls on consistent flows in dryland basins.

@@

The state of Arizona, located in the southwest United States, covers a total area of 295,253 km$^2$. Arizona is divided into two primary physiographic provinces: the Colorado Plateau in the northeast, and the Basin-and-Range region in the west and south. The Central Highlands is a transition zone consisting of scattered basins separated by the mountainous foothills of the Mogollon Rim. The Colorado Plateau is dominated by high-elevation desert with an average elevation of 1,936 masl (6,352 ft). Mean temperatures range from -6$^\circ$C (20$^\circ$F) to 26$^\circ$C (80$^\circ$F) and it averages 580 mm (23 in) of precipitation. The Basin-and-Range region has a semi-arid to arid climate with an average elevation of 490 masl (1600 ft). Average temperature ranges from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F) and the region averages 200 mm (8 in) of precipitation annually. ###CITATION###

@@

Arizona's hydrology varies seasonally and spatially between its physiographic regions. In the summer, localized and intense convective precipitation events are driven by the North American Monsoon, while in the winter, orographic precipitation comes from Pacific frontal systems [@eastoe2019mtnblock]. While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002]. As such, 94% of streams in Arizona are ephemeral or intermittent [@epa_AZephemeral]. Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack. While winter precipitation provides only 30% of annual averages, it provides the majority of water for natural reservoirs [@sheppard2002].

@@

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. @odonnell2016 found that 8-10 years of calibration data were necessary to account for climate variability in paired watershed studies in the region. Following this criteria, a minimum of 10 years of record was chosen. Any years missing more than 30 days of streamflow were removed from the possible record period. This analysis focuses on natural, streamflow-influencing recharge, excluding streamgages affected by regulation or diversions. These gages were identified through annual reports on water data published by the USGS [@usgs2010]. Finally, streamgages on the Colorado River were excluded from the dataset on the basis that they represent managed flows shared by Colorado River Compact states. Following these parameters, 205 USGS streamgages with acceptable periods of record were used in this study (@fig-study-area). Periods of record ranged from 10 to 112 years, with a median of 28 years.

Watersheds across the United States are delineated by the USGS using a hydrologically-defined network. This system delineates the country using hierarchical hydrologic unit codes (HUCs), where each subsequent basin includes the digits of the enclosing basin. Here, 8-digit HUCs (HUC 8s) are used to divide Arizona into 84 sub-basins that are fully or partially in the state (@fig-study-area). These HUC 8 sub-basins are analogous to medium-sized river basins and are defined by surface water characteristics.

Annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]). The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM]**.** Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance. Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].

@@

A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects. Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins. By calculating these percentages, we aimed to get a more comprehensive understanding of landscape composition across space. Land cover from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins. Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.

@@

Base flow was calculated using a single-parameter, recursive digital filter technique from [@nathan1990]. This base-flow separation technique is based on a recursive digital filter used in signal analysis that separates high-frequency signals (quickflow) from low-frequency signals (base flow) [@lyne1979]. @eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study. The Lyne-Hollick filter has been used in [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018]. It takes the form of

@@

Statisical analyses were performed on annual BFI and base flow values of instrumented streamgages. We examined temporal trends using the Mann-Kendall nonparametric trend test to determine the presence of monotonic trends [@kendall1970; @mann1945nonparametric]. The Mann-Kendall test was used to determine trends in BFI and trend significance of gauged reaches. This test is widely used in studies of this kind [@ficklin2016; @ayers2019; @woodhouse_udall22].

Trends in BFI and base flow were produced using a Mann-Kendall test for monotonic trends. This test expects that the data is non-parametric and that the observations are not autocorrelated over time. Observations at 4 streamgages show significant autocorrelation, though only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a significant trend in BFI. Due to this autocorrelation, the streamgage was removed from the trend analysis as the autocorrelation may inflate the variance of the Mann-Kendall statistic leading to an over/underestimate of the trend (CITATION).
